---
title: "Solutions for Lesson 4"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Function to read and tidy chick data
read_chick_data <- function(file){
  
  # Read the data in the file - starts at line 3
  chick_data <- read_delim(file, skip = 2, delim = " ", col_types = "ii")
  
  # Extract the diet from the file - first line of the file
  # and remove the word "Diet:" from it
  chick_diet <- read_lines(file, n_max = 1) %>% 
    str_remove("Diet:")

  # Get the chick ID - from the file name
  chick_id <- basename(file) %>% 
    str_remove(".txt") %>% str_remove("Chick_")
  
  # Add diet and chick ID to our table
  chick_data <- chick_data %>% 
    mutate(diet = chick_diet, id = chick_id)
  
  # Return the tidy chick_data
  return(chick_data)
}
```

These solutions use functions from the `tidyverse` meta-package.

```{r, message = FALSE}
# load the package
library(tidyverse)
```


## Understanding our data

If we open one of our files in a text editor, this is how they look like:

```{r, echo=FALSE}
cat(read_lines("data/chick_data/Chick_1.txt"), sep = "\n")
```

There are 3 pieces of information that we should bring together into a single 
tidy _data frame_:

* information about the diet applied - first line of the file 
* the weights measured on each day - third line of the file onwards, formated as a 
space-delimited table
* the ID of the chick - this is contained in the file name

That is, we should aim for something like this:

```{r, echo = FALSE}
read_chick_data("data/chick_data/Chick_1.txt") %>% knitr::kable()
```


## Building the steps to read and tidy our data

First do the task for a single file

```{r}
# Can read the table starting from line 3
read_delim("data/chick_data/Chick_1.txt", 
           delim  = " ", skip = 3)
```

etc... detail the other steps, which I've put together into a function below...


## Putting it all together into a function

We can wrap these steps into our own custom function. This will be convenient 
when we want to apply this function to the entire list of files.


```{r}
read_chick_data <- function(file){
  
  # Read the data in the file - starts at line 3
  chick_data <- read_delim(file, skip = 2, delim = " ", col_types = "ii")
  
  # Extract the diet from the file - first line of the file
  # and remove the word "Diet:" from it
  chick_diet <- read_lines(file, n_max = 1) %>% 
    str_remove("Diet:")

  # Get the chick ID - from the file name
  chick_id <- basename(file) %>% 
    str_remove(".txt") %>% str_remove("Chick_")
  
  # Add diet and chick ID to our table
  chick_data <- chick_data %>% 
    mutate(diet = chick_diet, id = chick_id)
  
  # Return the tidy chick_data
  return(chick_data)
}
```

Let's test the function:

```{r}
read_chick_data("data/chick_data/Chick_1.txt")
```



## Apply function to list of files

We can list the files in a directory:

```{r}
file_list <- list.files("data/chick_data/", pattern = ".txt",
                        full.names = TRUE)
head(file_list)
```

Then, we can apply a function to a list of values using the `map` function. 

```{r, eval=FALSE}
map(file_list, read_chick_data)
```

The `map()` function returns a list. This is sometimes convenient, but in our case
we want all of these tables to be bound together into a single `data.frame`. 

There are several variants of `map()`, which are designed for when we want the 
output to be of a defined type. In this case, we can use `map_df()`, which returns 
a `data.frame` instead of a `list`:

```{r}
# Apply our function to all of the files, and return a data.frame
chicks <- map_df(file_list, read_chick_data)

head(chicks)
```


## Analysing our data

The rest of the exercise is now applying what we've already learned in previous 
lessons:

#### Question 3

```{r}
# How many chicks survived the whole experiment?
## We can count the number of observations on each day
chicks %>% 
  count(Day)
```

Get IDs of surviving IDs

```{r}
# Get IDs of individuals that survived up to day 21
surviving_ids <- chicks %>% 
  group_by(id) %>% 
  filter(Day == 21) %>% 
  pull(id)  # this extracts the "id" column from the table and returns a vector

surviving_ids
```

Filter original table

```{r}
# Filter the table using this variable
my_chicks_survive <- chicks %>% 
  filter(id %in% surviving_ids)
```


#### Question 4a

```{r}
my_chicks_survive %>% 
  group_by(id) %>% 
  summarise(weight_dif = Weight[Day == 10] - Weight[Day == 0]) %>% 
  summarise(mean_weight_dif = mean(weight_dif))
```


#### Question 4b

```{r}
my_chicks_survive %>% 
  group_by(id, diet) %>% 
  summarise(weight_dif = Weight[Day == 10] - Weight[Day == 0]) %>% 
  group_by(diet) %>% 
  summarise(mean_weight_dif = mean(weight_dif))
```

#### Question 4c

```{r}
my_chicks_survive %>% 
  group_by(id, diet) %>% 
  summarise(weight_dif = Weight[Day == 10] - Weight[Day == 0]) %>% 
  ggplot(aes(diet, weight_dif)) + geom_boxplot()
```


#### Question 4d

```{r}
my_chicks_survive %>% 
  filter(Weight > 130) %>% 
  group_by(id) %>% 
  filter(Day == min(Day)) %>% 
  ggplot(aes(Day)) + geom_bar()
```


#### Question 4e

```{r}
ggplot(chicks, aes(Day, Weight, group = id, colour = diet)) +
  geom_line()
```


#### Question 4f

```{r}
chicks %>% 
  group_by(Day, diet) %>% 
  summarise(mean_weight = mean(Weight)) %>% 
  ggplot(aes(Day, mean_weight, colour = diet)) +
  geom_line()
```

Or:

```{r}
ggplot(chicks, aes(Day, Weight)) +
  geom_line(aes(group = id), colour = "grey") +
  geom_line(aes(colour = diet), stat = "summary", fun.y = "mean", size = 1)
```


#### Question 4g

...

